---
title: "Two Model solution"
output: html_notebook
---

## Model encoding

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(sf)
library(cetcolor)
# library(terra)
library(INLA)
library(spdep)
```

```{r}
data = readRDS(here('data/Rdata-clean.rds'))
gdf_traces = readRDS(here('data/Rdata-gdf_trace.rds'))
```

While I am testing/developing things, I'm only using a small subset of the data (around 1/2 a percent).
These are spatially distributed across the full region but further limited to the time period 1979--2010.
This still gives a few thousand observations to work with though.
Also for initial testing of everything, I am only including select time-invariant covariates.
Once I have a fully fleshed-out workflow though, I intend to add a few more, including some time-varying climate estimates (e.g. wind vectors).
I center the covariates about the mean and scale them with the standard deviation, and also normalize the discrete time dimension, setting the min year as Year 1.
I also perform a training/validation split for later use.

### Data filtering

```{r}
yr.start = 1985
yr.end = 2004
set.seed(777)
gdf_traces = gdf_traces %>% sample_n(100)
data = data %>% filter(trace_ID %in% gdf_traces$trace_ID) %>% 
  filter(Year >= yr.start) %>% filter(Year <= yr.end) %>% 
  group_by(trace_ID) %>% filter(all(yr.start:yr.end %in% Year)) %>% ungroup() %>% 
  arrange(trace_ID, Year)

# Select variables of interest
dat = data %>% select(trace_ID, East, North, Year, accum, 
                      std, elev, dem, slope, aspect, u10m, v10m, 
                      mu.u10, mu.v10)

# Center covariates
yr.min = min(data$Year)-1
dat = dat %>% 
  mutate(Year = Year-yr.min, elev = elev-mean(elev, na.rm=TRUE), 
         dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE), 
         u10m=u10m-mean(u10m, na.rm=TRUE), v10m=v10m-mean(v10m, na.rm=TRUE), 
         mu.u10=mu.u10-mean(mu.u10, na.rm=TRUE), mu.v10=mu.v10-mean(mu.v10, na.rm=TRUE)) #%>% 
  # mutate(Year = Year-mean(Year))

# Scale covariates by sd
dat = dat %>% 
  mutate(Year = Year, elev = elev/sd(elev), dem=dem/sd(dem), 
         slope=slope/sd(slope), u10m=u10m/sd(u10m), v10m=v10m/sd(v10m), 
         mu.u10=sd(mu.u10), mu.v10=sd(mu.v10))

# Split into training and testing sets
dat = dat %>% mutate(row.ID = 1:nrow(dat)) %>% relocate(row.ID)
dat.train = dat %>% slice_sample(prop = 0.80) %>% arrange(row.ID)
dat.test = dat %>% filter(!row.ID %in% dat.train$row.ID)
```

### Model 1

```{r}
mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
                    max.edge = c(25000, 75000), cutoff = 5000)
plot(mesh)
points(dat.train %>% select(East, North), col = "red")

spde = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(15000, 0.01), # P(range < 15 km) = 0.01
  prior.sigma = c(10, 0.05))

# Index for SPDE model and projector matrix
spat.idx = inla.spde.make.index(name = "spat.idx", n.spde = spde$n.spde)
A.spat = inla.spde.make.A(mesh = mesh, loc = cbind(dat.train$East, dat.train$North))
```

```{r}
# Make data stack
dat.stack <- inla.stack(
  data = list(y = dat.train$accum), 
  A = list(1, 1, 1, A.spat),
  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
                        slope=dat.train$slope, aspect=dat.train$aspect, 
                        u10m=dat.train$u10m, v10m=dat.train$v10m, 
                        mu.u10=dat.train$mu.u10, mu.v10=dat.train$mu.v10),
                 list(time = dat.train$Year),
                 spat.idx),
  tag = 'dat')

# Priors for temporal autocorrelation, fixed effects (currently empty), and gamma precision
time.spec = list(rho = list(prior = 'pc.cor1', param = c(0.3, 0.95)))
prior.fixed = list()
prec.prior = list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10%

# Model formula
form = y ~ -1 + Intercept + #modeling the intercept
  dem + aspect + #fixed effects
  f(spat.idx, model = spde) + #spatial random effect
  f(time, model = 'ar1', hyper = time.spec) #Temporal random effect modeled as AR1
```

```{r}
mod.1 = inla(form, 
               data = inla.stack.data(dat.stack), 
               family = 'Gamma',
               quantiles = NULL, 
               control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link = 1), 
               # control.fixed = prior.fixed, 
               control.family = list(hyper = list(prec = prec.prior)), 
               control.compute = list(waic=TRUE, return.marginals.predictor=TRUE, config=TRUE))

```

#### Posterior plots

Below is just a helper function for extracting and transforming marginal distributions to more intuitive scales.
I've then included some plots of the transformed marginal posteriors.
The hyperparameters for precisions and the sd's of spatial indices are left untransformed in these plots.
```{r}

t.marg = function(mod.marg, t.func, param.names=NULL, n=100) {
  
  if (any(class(mod.marg) == "matrix")) {
    mod.marg = list(mod.marg)
    if (!is.null(param.names)) {
      names(mod.marg = param.names)
    } else {
      names(mod.marg) = "UNNAMED"
    }
  }
  
  if (is.null(param.names)) {
    param.names = names(mod.marg)
  }
  
    if (is.list(t.func) != TRUE) {
    t.func = list(t.func)
  }
  if (length(t.func) == 1) {
    t.func = rep(t.func, length(param.names))
  }
  
  # Preallocate output tbl
  out = tibble(x=numeric(), y=numeric(), Param=vector(mode="character"))
  
  marg.names = names(mod.marg)
  
  for (i in 1:length(marg.names)) {
    marg.i = inla.tmarginal(t.func[[i]], mod.marg[[marg.names[i]]], n=n)
    out = out %>% bind_rows(tibble(x=marg.i[,1], y=marg.i[,2], Param=param.names[i]))
  }
  return(out)
}
```

```{r}
marg.fixed = t.marg(mod.1$marginals.fixed, function(x) exp(x))
HP.names = c("Gamma precision", "Range for spat.idx", "Stdev for spat.idx", 
             "Time precision", "Rho for time")
marg.hyper = t.marg(mod.1$marginals.hyperpar, function(x) (x), param.names = HP.names)
p.selection = c("Intercept", "Year", "dem", "aspect", "Gamma precision", "Range for spat.idx", 
                "Stdev for spat.idx", "Time precision", "Rho for time", "Stdev for spYR.idx", 
                "Range for spYR.idx")
tmp = marg.fixed %>% bind_rows(marg.hyper) %>% filter(Param %in% p.selection)
```

```{r fig, fig.width=16}
ggplot(tmp, aes(x=x,y=y)) + geom_line() + 
  facet_wrap(vars(Param), scales = "free") + 
  theme(text = element_text(size = 20))  
```

Calculate the model residuals
```{r}
# Get indices of observed data
fitted.idx = inla.stack.index(dat.stack, 'dat')$data

dat.train = dat.train %>% mutate(accum.mod=mod.1$summary.fitted.values$mean[fitted.idx]) %>% 
  mutate(accum.res = accum-accum.mod)

Is=sample(unique(dat.train$trace_ID),3)
ggplot(dat.train %>% filter(trace_ID %in% Is), aes(Year, accum.res, group=trace_ID)) + 
  geom_line(aes(color=as.factor(trace_ID))) + geom_smooth(method = 'lm')
```

Setup of residuals model
```{r}
# Make data stack
stack.res <- inla.stack(
  data = list(y = dat.train$accum.res), 
  A = list(1, 1, A.spat),
  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
                 tibble(Year=dat.train$Year), 
                 spat.idx),
  tag = 'dat')

# Model formula
form = y ~ -1 + Intercept + #modeling the intercept
  f(spat.idx,Year, model = spde) #spatial random effect
```

```{r}
mod.2 = inla(form, 
               data = inla.stack.data(stack.res), 
               family = 'Gaussian',
               quantiles = NULL, 
               control.predictor = list(compute = TRUE, A = inla.stack.A(stack.res)), 
               # control.fixed = prior.fixed, 
               control.family = list(hyper = list(prec = prec.prior)), 
               control.compute = list(waic=TRUE, return.marginals.predictor=TRUE, config=TRUE))
```

